---
title: "Hedge pricing"
date: today
code-fold: true
---

## Calculate implied volatility

First we have to extract nifty data using `tidyquant` or `Rquantlib`.

```{r message=FALSE}

# Install necessary packages if not already installed

# install.packages("quantmod")
# install.packages("RQuantLib")


# Load necessary packages

library(data.table)
library(tidyquant)

# Fetch the Nifty 50 index data
nifty_data <- tq_get("^NSEI", from = "1980-01-01", to = Sys.Date())

# Convert the data to a data.table
nifty_data <- data.table(nifty_data)
```

### Get closing price

```{r}
# Extract the last closing price directly from nifty_data
closing_price <- tail(nifty_data$close, 1)
closing_price
```

### Implied Volatility calculation

Plugin the latest option price for a liquid strike price with a 4 month maturity to get a reasonably accurate implied volatility.

```{r}

library(quantmod)
library(RQuantLib)

# Define option parameters for a single observation
nifty_price<-23290.5
option_price <- 105  # Current market price of the option
strike_price <- 22500  # Strike price of the option
time_to_expiry <- 19/365  # Time to expiration in years
risk_free_rate <- 0.071   # Risk-free interest rate
volatility_guess <- 0.2  # Initial guess for the volatility

# Calculate implied volatility for a put option

vol<- EuropeanOptionImpliedVolatility(type = "put",value =option_price ,underlying = nifty_price,strike = strike_price,dividendYield = 0,riskFreeRate = risk_free_rate,maturity = time_to_expiry,volatility = .2)

vol
```

At present the **implied volatility is `r round(vol*100)`**

## Option pricing calculation

Aim for a strike **less than 10 delta.** Plugin the **implied volatility** and other data.

```{r}

closing_price<-closing_price
strike <- closing_price * 0.85  # Strike price of the option
time_to_expiry <- 210/365  # Time to expiration in years
risk_free_rate <- 0.071   # Risk-free interest rate


price<-EuropeanOption(type = "put",underlying = closing_price,strike = strike,dividendYield = 0,riskFreeRate = risk_free_rate,maturity = time_to_expiry,volatility = vol)

cat("Closing price of nifty:\n")

print(closing_price)

cat("Strike price:\n")

strike
price

```

Price of a **4 month 5 delta option** is `r round(price$value)` for the given strike price

## Take home message

::: callout-note
### Summary

Before buying hedge, plugin the market price, strike price, time to maturity, implied volatility and risk free rate to the BS formula and find out the fair value. Then only buy the hedge.
:::

```{r}
closing<-21884.9
strikerate <- 20000  # Strike price of the option
time<- 210/365  # Time to expiration in years
risk_free_rate <- 0.071   # Risk-free interest rate
vol<-vol # implied volatility

price<-EuropeanOption(type = "put",underlying = closing,strike = strikerate,dividendYield = 0,riskFreeRate = risk_free_rate,maturity = time,volatility = vol)
price
strikerate
```

## **Monetising the hedge**

**§a** passive - at the time of expiry.

**§b** Empirical multiples of Initial cost. 

**§a** is ideal for long time horizon with potential big drawdown happening once or twice. **§b** out performs **§a** if no such tail events occur. In such a situation, even no-hedge strategy performs better than one with a hedge. 

For the hedging to outperform a pure buy and hold strategy with out hedging, it should be monitised at 8- 10 times the initial premium or a catastrophy should happen. In any case, the proceedings should be used to buy further far OTM puts and rest may be invested in the crashing asset.
