---
title: "Basic Biology"
format: html
order: 1
editor: visual
---

<iframe data-testid="embed-iframe" style="border-radius:12px" src="https://open.spotify.com/embed/episode/6lNFeoqmLYJ4iGUJCuIVYY?utm_source=generator" width="100%" height="152" frameBorder="0" allowfullscreen allow="autoplay; clipboard-write; encrypted-media; fullscreen; picture-in-picture" loading="lazy">

</iframe>

## Skeletal Muscle

### Structural Classification

Skeletal muscles are structurally classified as **pennate** or **non-pennate**.

-   **Pennate muscles** are further divided into:
    -   **Unipennate**
    -   **Bipennate**
    -   **Multipennate**

This classification is based on the orientation of muscle fibers relative to the tendon.

### Internal Structure

Under microscopy, skeletal muscle shows several landmarks:

-   **Z line**
-   **M line**
-   **H zone**
-   **A band**
-   **I band**

These structures are essential for understanding the **contractile mechanics** of muscle fibers.\
In electron microscopy, alternating **I** and **A bands** are visible.\
- **Thick filaments** contain *myosin*.\
- **Thin filaments** contain *actin*.

Together, myosin and actin form the basis of the **sliding filament mechanism** of muscle contraction.

------------------------------------------------------------------------

## Muscle Contraction

According to the **Sliding Filament Theory**, contraction begins when a **nerve impulse** reaches the **neuromuscular junction**.

1.  **Acetylcholine (ACh)** is released into the synapse.\
2.  It binds to receptors on the muscle membrane.\
3.  This triggers **calcium release** from the **sarcoplasmic reticulum**.\
4.  Calcium binds to **troponin**, shifting **tropomyosin** and exposing **actin binding sites**.\
5.  **Myosin heads** attach to actin, forming **cross-bridges** using **ATP**.

During contraction: - The **sarcomere shortens**.\
- The **A band** remains constant.\
- The **I band** and **H zone** narrow as actin and myosin overlap.

------------------------------------------------------------------------

## Energetics of Muscle

ATP is the **primary energy source** for muscle contraction.\
The body generates ATP through three energy systems:

| System | Duration | Description |
|------------------|-------------------------|-----------------------------|
| **ATP-CP (Creatine Phosphate)** | Up to 20 seconds | Immediate energy source |
| **Anaerobic (Lactic Acid) System** | 20–120 seconds | Moderate duration, no oxygen required |
| **Aerobic System** | Beyond 2 minutes | Sustained activity using oxygen |

------------------------------------------------------------------------

## Muscle Fiber Types

Muscle fibers differ in contraction speed, fatigue resistance, and metabolism.

| Fiber Type | Color | Speed | Fatigue Resistance | Metabolism | Motor Unit Size |
|------------|------------|------------|------------|------------|------------|
| **Type I (Slow-twitch, Red)** | Red | Slow | High | Aerobic | Small |
| **Type IIA (Fast-twitch Oxidative)** | Pink | Fast | Moderate | Aerobic & Anaerobic | Medium |
| **Type IIB (Fast Glycolytic)** | White | Very Fast | Low | Anaerobic | Large |

**Recruitment order:** Type II fibers are activated first during high-intensity effort, followed by Type I fibers for endurance.

------------------------------------------------------------------------

## Muscle Contractions

Muscle contractions are classified as:

1.  **Isometric** – Tension without change in length.\
2.  **Isotonic** – Change in length:
    -   **Concentric:** Muscle shortens.\
    -   **Eccentric:** Muscle lengthens.\
3.  **Isokinetic** – Constant speed contraction (using special equipment).

------------------------------------------------------------------------

## Peripheral Nerve Structure and Function

Peripheral nerves contain **afferent (sensory)**, **efferent (motor)**, and **sympathetic** fibers.

### Structure

-   **Myelinated and unmyelinated fibers** are present.\
-   **Nodes of Ranvier** facilitate **saltatory conduction** and **nutrient exchange**.

### Fiber Classification

| Fiber Type | Diameter | Myelination  | Speed    | Function              |
|------------|----------|--------------|----------|-----------------------|
| **Type A** | Large    | Heavy        | Fast     | Touch, proprioception |
| **Type B** | Medium   | Intermediate | Moderate | Autonomic             |
| **Type C** | Small    | None         | Slow     | Pain                  |

------------------------------------------------------------------------

## Resting and Action Potentials

-   **Resting Membrane Potential:** \~ -70 mV
    -   Maintained by ionic gradients and active transport.

### Action Potential Sequence

1.  **Depolarization:** Sodium channels open → Na⁺ influx.\
2.  **Threshold:** Around -55 mV.\
3.  **Repolarization:** Potassium channels open → K⁺ efflux.\
4.  **After-hyperpolarization:** Membrane becomes more negative before returning to baseline.

------------------------------------------------------------------------

## Nerve Injuries

Nerve injuries occur due to **compression, stretch, or laceration**.

### Classification

| Type | Description | Recovery |
|------------------|------------------------------|------------------------|
| **Neuropraxia (Sunderland I)** | Temporary conduction block | Full recovery |
| **Axonotmesis (Sunderland II)** | Axonal disruption, sheath intact | Good recovery |
| **Sunderland III–V** | Progressive sheath damage | Poor recovery |
| **Neurotmesis (Sunderland V)** | Complete transection | Poor prognosis |

------------------------------------------------------------------------

## Degeneration and Regeneration

### Wallerian Degeneration

-   Begins within **24 hours** after injury.\
-   Distal axon and myelin break down.\
-   **Macrophages** clear debris.

### Regeneration

-   **Schwann cells** proliferate and form **Bands of Büngner**.\
-   Axonal sprouts grow at **\~1 mm/day**.\
-   Success depends on alignment and trophic support.

**Delayed reinnervation** (after 12–18 months) leads to **irreversible muscle atrophy**.

------------------------------------------------------------------------

## Nerve Conduction Studies (NCS)

NCS evaluate peripheral nerve function by measuring:

-   **Sensory Nerve Action Potential (SNAP)**
-   **Compound Motor Action Potential (CMAP)**

| Parameter                 | Normal Value | Interpretation             |
|---------------------------|--------------|----------------------------|
| **Velocity (Upper limb)** | \> 45 m/s    | ↓ velocity → Demyelination |
| **Velocity (Lower limb)** | \> 35 m/s    | ↑ latency → Demyelination  |
| **Latency (Wrist)**       | \< 3.5 ms    | ↓ amplitude → Axonal loss  |

------------------------------------------------------------------------

## Electromyography (EMG)

Used to study **electrical activity** of skeletal muscle, typically **3 weeks post-injury**.

### Key Findings

-   **Positive sharp waves**
-   **Fibrillations**
-   **Fasciculations**
-   **Myokymic discharges**

**Fibrillations** indicate **axonotmesis or neurotmesis**.\
Absence of fibrillations suggests **neuropraxia**.

------------------------------------------------------------------------

## Tendons and Ligaments

### Tendons

-   Connect **muscle to bone**.\
-   May be **sheathed** (with vincula) or covered by **paratenon**.\
-   Composed mainly of **Type I collagen**, with some **Type III collagen** and **elastin**.\
-   **Fibroblasts** are the main cells.

### Ligaments

-   Connect **bone to bone**.\
-   Covered by **epiligament**.\
-   Contain **more elastin** and **less uniform fiber orientation** than tendons.

------------------------------------------------------------------------

## Biomechanical Properties

| Tissue                  | Tolerable Strain Before Failure |
|-------------------------|---------------------------------|
| **Tendons & Ligaments** | 5–10%                           |
| **Bone**                | 1–4%                            |

Their elasticity supports **dynamic musculoskeletal movement**.

------------------------------------------------------------------------

## Healing of Tendons

Tendon healing occurs in **three overlapping phases**:

| Phase | Duration | Key Events |
|------------------|-------------------------|-----------------------------|
| **Inflammation** | 0–5 days | Hematoma, inflammatory cell recruitment |
| **Proliferation** | 3 days – 6 weeks | Fibroblast proliferation, collagen bridge formation |
| **Remodeling** | 6 weeks – 1 year | Collagen maturation and alignment |

After surgical repair: - **Weakest point:** 1 week\
- **Significant strength gain:** 4 weeks\
- **Near-normal strength:** 6 months

------------------------------------------------------------------------

## Self Assessment

```{r echo=FALSE, warning=FALSE}
library(exams)
library(exams2forms)

exm <- list("_exams/ex1.Rmd", "_exams/ex2.Rmd", "_exams/ex3.Rmd", "_exams/ex4.Rmd", "_exams/ex5.Rmd", "_exams/ex6.Rmd", "_exams/ex7.Rmd", "_exams/ex8.Rmd")

```

```{r echo = FALSE, message = FALSE, results="asis"}
exams2forms(exm)
```
